---
title: "ST 695 Sequential Split Plot Designs"
output: word_document
date: "2025-05-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code chunk produces a design matrix for a split plot design given the number of HTC Factors (HTC), Easy to Change Factors (ETC), and replicates for HTC Factors 
```{r }

HTC <- 2
ETC <- 2
r <- 2
nsim <- 100

# Create the factors for the Design Matrix
w_names <- paste0("w", 1:HTC)
s_names <- paste0("s", 1:ETC)

# Create list of w and s variables
combined_list <- c(
  setNames(replicate(ETC, c(-1, 1), simplify = FALSE), s_names),
  setNames(replicate(HTC, c(-1, 1), simplify = FALSE), w_names)
)

# Create data frame of all combinations of w and s
df <- expand.grid(combined_list) #base experimental design
df_rep <- df[rep(seq_len(nrow(df)), r), ] #replicated experimental design
row.names(df_rep) <- NULL #make the row count sequential

# Build the formula for linear and interaction terms (no quadratic)
linear_terms <- c(w_names, s_names)
fo <- as.formula(
  paste("~ (", paste(linear_terms, collapse = " + "), ")^2")
)

# Create the model matrix
X <- model.matrix(fo, data = df_rep)
colnames(X)[1] <- '0'

```


Simulate response data by specifying $\beta_r, \beta_{rs}, \sigma^2_B, \sigma^2$. The inputs to the simulation are Beta, sigma2_B, sigma2. Repeat this process 100 times, each time you generate a Y, fit the model, and save the estimates. At the end you should have 100 estimates for the intercept, 100 for w1, etc. Then with the estimates you can plot a histogram of each one and they should all be centered at the true values (except for the variance histograms; those will be right skewed).

```{r}

library(lme4)

set.seed(1)
  dl <- as.factor(rep(seq(1, r*2^HTC), each = 2^ETC))  
    
  fixed_effect_names <- paste0("beta", colnames(X))
  
  beta <- setNames(rnorm(length(fixed_effect_names)),
  fixed_effect_names)  #assign values to fixed effect names


#Simulation Function  
sim_data <- function(X, beta, sigma2, sigma2_B, dl) {
  
  mu <- X %*% beta
  n_groups <- length(unique(dl))
  
  
#  T <- data.frame(replicate(runs,rnorm(n=n,mean=m,sd=sqrt(v))))
    
  delta <- rnorm(n_groups, mean = 0, sd = sqrt(sigma2_B))
  delta <- rep(delta, times = table(dl))
  
  epsilon <- rnorm(nrow(X), mean = 0, sd = sqrt(sigma2))
  
  Y <- mu + delta + epsilon
  
  df <- data.frame(delta, df_rep, Y)
  
  fo <- as.formula(
  paste("Y ~ (", paste(linear_terms, collapse = " + "), ")^2 + (1|delta)"))
  
  rmodel <- lmer(fo, data = df)
  
  # Extract fixed effects estimates
  
  fixed_effects <- as.list(fixef(rmodel))

# Extract variance-covariance matrix of fixed effects
  random_effects <- VarCorr(rmodel)

  varcomp <- as.data.frame(random_effects)

  sigma2_B <- varcomp$vcov[1]
  sigma2 <- varcomp$vcov[2]
  
  return(list(feff = fixed_effects, sigma2_B = sigma2_B, sigma2 = sigma2))

}

# Set up empty data frame

sim_df <- data.frame(matrix(ncol = length(beta) + 2, nrow = 0))

colnames(sim_df) <- c(colnames(X),"sigma2_B", "sigma2")

#Run Simulation
for(i in 1:nsim){
 
  rmodel <- as.data.frame(sim_data(X, beta, sigma2 = 1, sigma2_B = 1, dl))
  
  sim_df <- rbind(sim_df, rmodel)

}

for (col in 1:ncol(sim_df)) {
    hist(sim_df[,col], main=paste("Histogram of ", colnames(sim_df)[col]),  xlab=colnames(sim_df)[col])
}



```

Simulation Summary Results
```{r}



```