---
title: "ST 695 Sequential Split Plot Design SW Model"
output: word_document
date: "2025-06-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code chunk produces a design matrix for a split plot design given the number of HTC Factors (HTC), Easy to Change Factors (ETC), and replicates for HTC Factors.  This model includes HTC and ETC Factors. The model of interest here is given by

$y_{ikl}^{S} = \beta_0^* + \sum_{p=1}^{w} \beta_p x_{pi} + e_i^{W} + \sum_{q=1}^{s} \gamma_q z_{ikq} + e_{ik}^{S} + e_{ikl}^{SS}$
```{r }

################################################################################
#Create the model matrix for X_S (Whole Plot and Split Plot Factors)
################################################################################

HTC <- 2
ETC <- 2
r <- 2
nsim <- 100

# Create the factors for the Design Matrix
w_names <- paste0("w", 1:HTC)
s_names <- paste0("s", 1:ETC)

# Create list of w and s variables
combined_list <- c(
  setNames(replicate(ETC, c(-1, 1), simplify = FALSE), s_names),
  setNames(replicate(HTC, c(-1, 1), simplify = FALSE), w_names)
)

# Create data frame of all combinations of w and s
df_S <- expand.grid(combined_list) #base experimental design
df_rep_S <- df_S[rep(seq_len(nrow(df_S)), r), ] #replicated experimental design
row.names(df_rep_S) <- NULL #make the row count sequential

# Build the formula for linear and interaction terms (no quadratic)
linear_terms_ws <- c(w_names, s_names)
formula_S <- as.formula(
  paste("~ (", paste(linear_terms_ws, collapse = " + "), ")^2")
)

# Create the model matrix for X_S (Split and Whole Plot Factors)
X_S <- model.matrix(formula_S, data = df_rep_S)
colnames(X_S)[1] <- '0'

################################################################################
#Create the model matrix for X_W (Whole Plot Factors)
################################################################################

#1. Create list of whole plot variables
wpf_terms <- setNames(replicate(HTC, c(-1,1), simplify = FALSE), w_names)


#2. Create data frame of all combinations of w
df_W <- expand.grid(wpf_terms) #base experimental design
df_rep_W <- df_W[rep(seq_len(nrow(df_W)), r), ] #replicated experimental design
row.names(df_rep_W) <- NULL #make the row count sequential

#3. Build the formula for linear and interaction terms (no quadratic)
linear_terms_w <- w_names

formula_W <- as.formula(
  paste("~ (", paste(w_names, collapse = " + "), ")^2")
)

# Create the model matrix
X_W <- X_S[, !grepl("s", colnames(X_S), ignore.case = TRUE)]

################################################################################
#Create the model matrix for X_SW (Y_S - Y_W Model)
################################################################################

#Remove columns with Whole Plot Terms only (Main Effects and Interaction of WPF)
X_SW <- X_S[, grep('s', colnames(X_S), value = TRUE)]
colnames(X_SW) <- gsub(":", "_", colnames(X_SW))

```


Simulate response data. The inputs to the simulation are Beta, sigma2_W, sigma2_S. Repeat this process 100 times, each time you generate a Y, fit the model, and save the estimates.Plot a histogram of each estimate and they should all be centered at the true values (except for the variance histograms; those will be right skewed).  This is the S Model.

The W Model is then analyzed (Whole Plot Factors only) and finally the S_W Model which is Y_S - Y_W.

```{r}

library(lmerTest)

sigma2_W <- 0.2
sigma2_S <- 1

set.seed(1)
  dl <- as.factor(rep(seq(1, r*2^HTC), each = 2^ETC))  #delta as a factor
    
  fixed_effect_names_S <- paste0("beta", colnames(X_S))
  fixed_effect_names_W <- paste0("beta", colnames(X_W))
  
  #beta <- setNames(rnorm(length(fixed_effect_names))
  # fixed_effect_names)  #assign values to fixed effect names
  effects <- c(10,2.5,-1,1.5,0,0,1/2,0,0,0,0)
  
  #assign values to fixed effect names
  beta_S <- setNames(effects, fixed_effect_names_S)  
  names(beta_S)[1] <- "beta*0"
  beta_W <- beta_S[!grepl("s", colnames(X_S), ignore.case = TRUE)]
 
  
  #Compute the means of S and W model
  mu_S <- X_S %*% beta_S
  mu_W <- X_W %*% beta_W

#Simulation Function for S Model
sim_model_S <- function(X_S, mu_S, mu_W, sigma2_S, sigma2_W, dl) {
  
  n_groups <- length(unique(dl))
  
  epsilon_W <- rnorm(n_groups, mean = 0, sd = sqrt(sigma2_W))
  epsilon_W <- rep(epsilon_W, times = table(dl))
  
  epsilon_S <- rnorm(nrow(X_S), mean = 0, sd = sqrt(sigma2_S))
  
  Y_S <- mu_S + epsilon_W + epsilon_S
  Y_W <- mu_W + epsilon_W

  df_S <- data.frame(epsilon_W, df_rep_S, Y_S)
  
  formula_S <- as.formula(
  paste("Y_S ~ (", paste(linear_terms_ws, collapse = " + "), ")^2 + (1|epsilon_W)"))
  
  rmodel_S <- summary(lmer(formula_S, data = df_S))
  rownames(rmodel_S$coefficients) <- sub("\\(Intercept\\)", "Intercept", rownames(rmodel_S$coefficients))

  # Extract fixed effects estimates

  fixed_effects_S <- rmodel_S$coefficients[, "Estimate"]
  pvalues_S <- rmodel_S$coefficients[, "Pr(>|t|)"]

   
  # Extract and compute variance of random effects
  
  sigma2_W <- (as.numeric(rmodel_S$varcor$epsilon_W))
  sigma2_S <- (rmodel_S$sigma)^2
 
   # Generate data frame of fixed and variable effects
  dffe_S <- as.data.frame(t(fixed_effects_S))
  colnames(dffe_S) <- paste0("Fixed_Eff_Model_S (", names(fixed_effects_S), ")")
  dffp_S <- as.data.frame(t(pvalues_S))
  colnames(dffp_S) <- paste0("pvalue_Model_S (", names(pvalues_S), ")")
  df_out_S <- cbind(dffe_S, dffp_S)
  df_out_S$sigma2_W <- sigma2_W
  df_out_S$sigma2_S <- sigma2_S
  return(list(df_out_S = df_out_S, Y_W = Y_W, Y_S = Y_S))
  
}

#Simulation Function for W Model
sim_model_W <- function(Y_W) {
  
  Y_W_unique <- unique(Y_W)  
  df_W <- data.frame(df_rep_W, Y_W = Y_W_unique)

  # Build the formula for linear and interaction terms (no quadratic)
  linear_terms <- w_names

  formula_W <- as.formula(
  paste("Y_W ~ (", paste(w_names, collapse = " + "), ")^2"))
  
  rmodel_W <- summary(lm(formula_W, data = df_W))
  rownames(rmodel_W$coefficients) <- sub("\\(Intercept\\)", "Intercept", rownames(rmodel_W$coefficients))

  # Extract fixed effects estimates
  fixed_effects_W <- rmodel_W$coefficients[, "Estimate"]
  pvalues_W <- rmodel_W$coefficients[, "Pr(>|t|)"]

  # Extract and compute variance of random effects
   sigma2_W <- rmodel_W$sigma
  
  
  # Combine into a data frame 
  dffe_W <- as.data.frame(t(fixed_effects_W))
  colnames(dffe_W) <- paste0("Fixed_Eff_Model_W (", names(fixed_effects_W),")")
  dffp_W <- as.data.frame(t(pvalues_W))
  colnames(dffp_W) <- paste0("pvalue_Model_W (", names(pvalues_W), ")")
  df_out_W <- cbind(dffe_W, dffp_W)
  df_out_W$sigma2_W <- sigma2_W
  return(df_out_W)
  
}

#Simulation Function for SW Model
sim_model_SW <- function(X_SW, Y_S, Y_W) {

  df_SW <- data.frame(Y_SW = Y_S - Y_W, X_SW)

# Build the formula for linear and interaction terms (no quadratic)

 formula_SW <- as.formula(
  paste("Y_SW ~", paste(colnames(X_SW), collapse = " + "))
)

  rmodel_SW <- summary(lm(formula_SW, data = df_SW))
  rownames(rmodel_SW$coefficients) <- sub("\\(Intercept\\)", "Intercept", rownames(rmodel_SW$coefficients))

  # Extract fixed effects estimates

  fixed_effects_SW <- rmodel_SW$coefficients[, "Estimate"]
  pvalues_SW <- rmodel_SW$coefficients[, "Pr(>|t|)"]

  # Extract and compute variance of random effects
   sigma2_S <- rmodel_SW$sigma

   # Combine into a data frame
  dffe_SW <- as.data.frame(t(fixed_effects_SW))
  colnames(dffe_SW) <- paste0("Fixed_Eff_Model_SW (", names(fixed_effects_SW),")")
  dffp_SW <- as.data.frame(t(pvalues_SW))
  colnames(dffp_SW) <- paste0("pvalue_Model_SW (", names(pvalues_SW), ")")
  df_out_SW <- cbind(dffe_SW, dffp_SW)
  df_out_SW$sigma2_S <- sigma2_S
  return(df_out_SW)

}

#Analysis Function
analyze <- function(sim_df) {
  
  fixed_eff_cols <- grep("^Fixed_Eff", colnames(sim_df), value = TRUE)

  for (col in fixed_eff_cols) {
    hist(sim_df[[col]], main = paste("Histogram of", col), xlab = col)
}


#Plot Histograms of Variance Components
varcomp_cols <- grep("^sigma", colnames(sim_df), value = TRUE)
  for (col in varcomp_cols){
    hist(sim_df[[col]], main = paste("Histogram of", col), xlab = col)
}

#Statistical Significance
# Get the names of p-value columns
pval_cols <- grep("^pvalue", colnames(sim_df), value = TRUE)

# Create a logical matrix: TRUE if p-value < 0.05
sig_matrix <- sim_df[, pval_cols] < 0.05

#rename the columns of sig matrix TPR.
colnames(sig_matrix) <- gsub("pvalue", "TPR", colnames(sig_matrix))

# Compute proportion significant for each estimate
prop_sig <- colMeans(sig_matrix, na.rm = TRUE)

# Print results
print(prop_sig)
  
}

# Set up empty data frames for simulation results

sim_df_S <- data.frame()
sim_df_W <- data.frame()
sim_df_S_W <- data.frame()
  
#Run Simulation
for(i in 1:nsim){
 
  sim_model_S_out <- sim_model_S(X_S, mu_S, mu_W, sigma2_S = sigma2_S, sigma2_W = sigma2_W, dl)
  Y_W <- sim_model_S_out$Y_W
  Y_S <- sim_model_S_out$Y_S
  
  rmodel_S <- as.data.frame(sim_model_S_out$df_out_S)
  sim_df_S <- rbind(sim_df_S, rmodel_S)
  
  rmodel_W <- as.data.frame(sim_model_W(Y_W))
  sim_df_W <- rbind(sim_df_W, rmodel_W)
  
  rmodel_S_W <- as.data.frame(sim_model_SW(X_SW, Y_S, Y_W))
  sim_df_S_W <- rbind(sim_df_S_W, rmodel_S_W)
 
}

S_model_analysis <- analyze(sim_df_S)
W_model_analysis <- analyze(sim_df_W)
SW_model_analysis <- analyze(sim_df_S_W)


```

