---
title: "ST 695 Split Plot Designs Split Plot Model"
output: word_document
date: "2025-06-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code chunk produces a design matrix for a split plot design where the main effects (Whole Plot Factors) have been removed.

$y_{ikl}^{S} - y_{ij}^{W}  = (\beta_0^* - \beta_0) + \sum_{q=1}^{s}\gamma_q z_{ikq} + e_{ik}^{S} + e_{ikl}^{SS} -e_{ij}^{WS} $
```{r }

HTC <- 2
ETC <- 2
r <- 2
nsim <- 100

# Create the factors for the Design Matrix
w_names <- paste0("w", 1:HTC)
s_names <- paste0("s", 1:ETC)

# Create list of w and s variables
combined_list <- c(
  setNames(replicate(ETC, c(-1, 1), simplify = FALSE), s_names),
  setNames(replicate(HTC, c(-1, 1), simplify = FALSE), w_names)
)

# Create data frame of all combinations of w and s
df <- expand.grid(combined_list) #base experimental design
df_rep <- df[rep(seq_len(nrow(df)), r), ] #replicated experimental design
row.names(df_rep) <- NULL #make the row count sequential

# Build the formula for linear and interaction terms (no quadratic)
linear_terms <- c(w_names, s_names)
fo <- as.formula(
  paste("~ (", paste(linear_terms, collapse = " + "), ")^2")
)

# Create the model matrix
X <- model.matrix(fo, data = df_rep)
colnames(X)[1] <- '0'


#remove columns with Whole Plot Terms only (Main Effects and Interaction of WPF)
X_reduced <- X[, grep('s|0', colnames(X), value = TRUE)]
linear_terms <- colnames(X_reduced)


```


Simulate response data. The inputs to the simulation are Beta, sigma2_W, sigma2_S. Repeat this process 100 times, each time you generate a Y, fit the model, and save the estimates. At the end you should have 100 estimates for the intercept, 100 for w1, etc. Then with the estimates you can plot a histogram of each one and they should all be centered at the true values (except for the variance histograms; those will be right skewed).

```{r}

sigma2_W <- 0.2
sigma2_S <- 1

set.seed(1)
  dl <- as.factor(rep(seq(1, r*2^HTC), each = 2^ETC))  
    
  fixed_effect_names <- paste0("beta", colnames(X_reduced))
  
  # beta <- setNames(rnorm(length(fixed_effect_names)),
  # fixed_effect_names)  #assign values to fixed effect names
  effects <- c(10,2.5,-1,1.5,0,0,1/2,0)
  beta <- setNames(effects, fixed_effect_names)  #assign values to fixed effect names
  

#Simulation Function  
sim_data <- function(X_reduced, beta, sigma2_S, sigma2_W) {
  
  mu <- X_reduced %*% beta
  n_groups <- length(unique(dl))
  
  
  delta <- rnorm(n_groups, mean = 0, sd = sqrt(sigma2_W))
  delta <- rep(delta, times = table(dl))
  
  epsilon <- rnorm(nrow(X_reduced), mean = 0, sd = sqrt(sigma2_S))
  
  Y <- mu + delta + epsilon
  
  df <- data.frame(delta, df_rep, Y)
  
  fo <- as.formula(
  paste("Y ~ (", paste(linear_terms, collapse = " + "), ")^2 + (1|delta)"))
  
  rmodel_p <- summary(lmer(fo, data = df))

  # Extract fixed effects estimates

 #   fixef_names <- rownames(rmodel_p$coefficients)
  fixed_effects <- rmodel_p$coefficients[, "Estimate"]
  pvalues <- rmodel_p$coefficients[, "Pr(>|t|)"]

   
 
# Extract and compute variance of random effects
  
  sigma2_W <- (as.numeric(rmodel_p$varcor$delta))
  sigma2_S <- (rmodel_p$sigma)^2
 
   # Combine into a data frame 
  dffe <- as.data.frame(t(fixed_effects))
  colnames(dffe) <- paste0("fixed_eff ", names(fixed_effects))
  dffp <- as.data.frame(t(pvalues))
  colnames(dffp) <- paste0("pvalue ", names(pvalues))
  df_out <- cbind(dffe, dffp)
  df_out$sigma2_W <- sigma2_W
  df_out$sigma2_S <- sigma2_S
  return(df_out)
  

}

# Set up empty data frame

sim_df <- data.frame()
  
#Run Simulation
for(i in 1:nsim){
 
  rmodel <- as.data.frame(sim_data(X, beta, sigma2_S = sigma2_S, sigma2_W = sigma2_W, dl))
  sim_df <- rbind(sim_df, rmodel)

}

#Plot Histograms of Fixed Effects
fixed_eff_cols <- grep("^fixed_eff", colnames(sim_df), value = TRUE)

  for (col in fixed_eff_cols) {
    hist(sim_df[[col]], main = paste("Histogram of", col), xlab = col)
}


#Plot Histograms of Variance Components
varcomp_cols <- grep("^sigma", colnames(sim_df), value = TRUE)
  for (col in varcomp_cols){
    hist(sim_df[[col]], main = paste("Histogram of", col), xlab = col)
}

#Statistical Significance
# Get the names of p-value columns
pval_cols <- grep("^pvalue", colnames(sim_df), value = TRUE)

# Create a logical matrix: TRUE if p-value < 0.05
sig_matrix <- sim_df[, pval_cols] < 0.05

#rename the columns of sig matrix TPR.
colnames(sig_matrix) <- gsub("pvalue", "TPR", colnames(sig_matrix))

# Compute proportion significant for each estimate
prop_sig <- colMeans(sig_matrix, na.rm = TRUE)

# Print results
print(prop_sig)


```

